---
title: "SST MHW Simulations - CBWMWT"
output:
  html_document:
    df_print: paged
---

July 2022

This analysis is focused on modeling effects of trends in SST mean and variance following Oliver 2019. The site is at the former location of NOAA NDBC CHLV2, east of the mouth of Chesapeake Bay. 

This block reads in data and calculates values needed for model (mean, variance, tau, se)

```{r}
library(lubridate)
XDdf=read.csv("SST_COC_CBW_Jul2022.csv") #meas WT avgd over LST
Xdf=XDdf[,c(6,11,16)] #CBWMWT WTDA, seas, anom
dt=mdy(XDdf$dt)
dd=decimal_date(dt)
Ddf=data.frame(dt,dd,year(dt),month(dt),day(dt))
colnames(Ddf)=c("Date","DecY","Y","M","D")
```

```{r}
LD=length(Ddf$Y)
ild=which(Ddf$M==2&Ddf$D==29)
if (length(ild)>0) Ddf=Ddf[-ild,]
YV=unique(Ddf$Y)
NY=length(YV)

NR=length(Ddf$Y)
LX=dim(Xdf)[1]
print(c(NY,NR,LX,LD))

if(LX==LD & length(ild>0)) Xdf=Xdf[-ild,]
if(LX!=LD & LX!=NR) print (c("input dataframes not same length"))
```

```{r}
library(heatwaveR)
Hdf=data.frame(Ddf$Date,Xdf[,1])
colnames(Hdf)=c("t","temp")
YHdf=Hdf; YHdf$Y=Ddf$Y; YHdf$DecY=Ddf$DecY
Xmean=aggregate(YHdf$temp ~ YHdf$Y, FUN=mean, na.action=na.pass)
colnames(Xmean)=c("Y","Tmean")
Xclim=ts2clm(Hdf,climatologyPeriod = c("1987-01-01", "2016-12-31"), clmOnly=TRUE)
Xthresh=Xclim$thresh[-60]  #ignore leapday
XthV=rep(Xthresh,NY)
Xseas=Xclim$seas[-60]  #ignore leapday
XsV=rep(Xseas,NY)
YHdf$tmseas=Hdf$temp-XsV
SSTAnom=YHdf$tmseas
fitmS=lm(YHdf$tmseas ~ YHdf$DecY)
trend=fitmS$coefficients[1]+fitmS$coefficients[2]*YHdf$DecY
Xtrend=fitmS$coefficients[2]
Xincpt=fitmS$coefficients[1]
Xtrci=confint(fitmS)[2,2]-fitmS$coefficients[2]
YHdf$tdT=YHdf$tmseas-trend #detrend
Xvar=aggregate(YHdf$tdT ~ YHdf$Y, FUN=var, na.action=na.pass)
colnames(Xvar)=c("Y","Tvar")
```

```{r}
Xhi=rep(0,NR); N28=Xhi
Xhi[(Hdf$temp-XthV)>0]=1
N28[Hdf$temp>=28]=1
a28=aggregate(N28 ~ Ddf$Y, FUN=sum, na.action=na.pass)    
colnames(a28)=c("Y","abv28")
Xa28=rep(0,length(YV))
Xa28[a28$Y-YV[1]+1]=a28$abv28
frachi=sum(Xhi)/NR
dX=diff(Xhi)
nd = which(dX==1)
np = which(dX==-1)
if (np[1]<nd[1]) np=np[-1]
if (length(np) != length(nd)) nd=nd[-length(nd)]
durXHi=np-nd
n5=which(durXHi>=5)

Dexp=aggregate(durXHi[n5] ~ Ddf$Y[nd[n5]], FUN=sum, na.action=na.pass)
Ddur=aggregate(durXHi[n5] ~ Ddf$Y[nd[n5]], FUN=mean, na.action=na.pass)
colnames(Dexp)=c("Y","expMHW")
colnames(Ddur)=c("Y","durMHW")
Xexp=rep(0,length(YV))
Xexp[Dexp$Y-YV[1]+1]=Dexp$expMHW
Xdur=rep(0,length(YV))
Xdur[Ddur$Y-YV[1]+1]=Ddur$durMHW

inten=rep(0,length(n5))
for (k in 1:length(n5)){
  inten[k]=max(SSTAnom[nd[n5[k]]:np[n5[k]]])
}
Dint=aggregate(inten ~ Ddf$Y[nd[n5]], FUN=max, na.action=na.pass)
colnames(Dint)=c("Y","intMHW")
Xint=rep(0,length(YV))
Xint[Dint$Y-YV[1]+1]=Dint$intMHW

Dmhw=aggregate(nd[n5] ~ Ddf$Y[nd[n5]], FUN=NROW, na.action=na.pass)
colnames(Dmhw)=c("Y","nMHW")
Xmhw=rep(0,length(YV))
Xmhw[Dmhw$Y-YV[1]+1]=Dmhw$nMHW
```


```{r}
print(frachi)
print(Xtrend)
```

```{r}
plot(YV,Xint)
fitint = lm(Xint ~ YV)
abline(fitint)
```

```{r}
plot(YV,Xexp)
fitexp = lm(Xexp ~ YV)
abline(fitexp)
```

```{r}
plot(YV,Xa28)
```

```{r}
plot(YV,Xdur)
fitdur = lm(Xdur ~ YV)
abline(fitdur)
```


```{r}
plot(YV,Xmhw)
fitmhw = lm(Xmhw ~ YV)
abline(fitmhw)
```

Save values for plotting
```{r}
M = rbind(Xexp,Xint,Xdur,Xa28,Xmhw)
#write.csv(M,"MHWCV.csv")
```

Calculate parameters for autoregressive model ala Oliver 2019

```{r}
fitL=lm(YHdf$tdT[2:NR]~YHdf$tdT[1:(NR-1)])
slope=as.numeric(fitL$coefficients[2])
sig_e=sd(fitL$residuals)          
tau=-1/log(slope)

print(c(slope,sig_e,tau))
```

Calculate 500 instances of SST time series using autoregressive model with parameters slope and se from above.

```{r}
NS=500  #number of simulations
NR=length(Ddf$Y)
YV=unique(Ddf$Y)
NY=length(YV)
Q=matrix(0,nrow=NR,ncol=NS)

for (j in 1:NS){
  Er = rnorm(NR,mean=0,sd=sig_e)
  for (i in 2:NR) {Q[i,j] = slope*Q[i-1,j] + Er[i]}
}
```

Add trend and seasonal cycle back into time series. 

```{r}
Qt=Q+trend  #add trend from SST data
Qst=Qt+XsV   #add seasonal cycle to trend
Qs=Q+XsV  #add seasonal cycle w/o trend
XY=Ddf$Y
```

```{r}
vQ=apply(Q,2,var)
summary(vQ)
var(SSTAnom)
```

Calculate exposure and intensity for simulated SST with trend included.

```{r}
Qexp=matrix(0,nrow=NY,ncol=NS); Qa28=Qexp; Qint=Qexp; Qdur=Qexp; Qmhw=Qexp
frachi = rep(0,NS); cnt = matrix(0, nrow=NS, ncol=50)
for (j in 1:NS) {
  Xhi=rep(0,NR); N28=Xhi
  Xhi[(Qst[,j]-XthV)>0]=1
  S1=rle(Xhi)$lengths; T1=rle(Xhi)$values
  U1=S1*T1;  E1=U1[-(which(U1==0))]
  evtab = table(factor(E1, levels=1:50))
  cnt[j,]=evtab
  frachi[j]=sum(Xhi)/NR
  dX=diff(Xhi)
  nd = which(dX==1)
  np = which(dX==-1)
  if (np[1]<nd[1]) np=np[-1]
  if (length(np) != length(nd)) nd=nd[-length(nd)]
  durXHi=np-nd
  n5=which(durXHi>=5)
  inten=rep(0,length(n5))
  for (k in 1:length(n5)){
    inten[k]=max(Qt[nd[n5[k]]:np[n5[k]],j])
  }
  Dint=aggregate(inten ~ XY[nd[n5]], FUN=max, na.action=na.pass)
  colnames(Dint)=c("Y","intMHW")
  Sint=rep(0,NY)
  Sint[Dint$Y-YV[1]+1]=Dint$intMHW
  Qint[,j]=Sint
  
  Dexp=aggregate(durXHi[n5] ~ XY[nd[n5]], FUN=sum, na.action=na.pass)
  colnames(Dexp)=c("Y","expMHW")
  Sexp=rep(0,NY)
  Sexp[Dexp$Y-YV[1]+1]=Dexp$expMHW
  Qexp[,j]=Sexp
  
  N28[Qst[,j]>=28]=1
  a28=aggregate(N28 ~ XY, FUN=sum, na.action=na.pass)    
  colnames(a28)=c("Y","abv28")
  Sa28=rep(0,NY)
  Sa28[a28$Y-YV[1]+1]=a28$abv28
  Qa28[,j]=Sa28
  
  Ddur=aggregate(durXHi[n5] ~ XY[nd[n5]], FUN=mean, na.action=na.pass)
  colnames(Ddur)=c("Y","durMHW")
  Sdur=rep(0,NY)
  Sdur[Ddur$Y-YV[1]+1]=Ddur$durMHW
  Qdur[,j]=Sdur
  
  Dmhw=aggregate(nd[n5] ~ XY[nd[n5]], FUN=NROW, na.action=na.pass)
  colnames(Dmhw)=c("Y","nMHW")
  Smhw=rep(0,NY)
  Smhw[Dmhw$Y-YV[1]+1]=Dmhw$nMHW
  Qmhw[,j]=Smhw
}
```


```{r}
boxplot(t(Qint),outpch=20, outcex=0.5)
points(Xint,col="blue")
```
```{r}
boxplot(t(Qexp),outpch=20, outcex=0.5)
points(Xexp,col="red")
```


```{r}
boxplot(t(Qa28),outpch=20, outcex=0.5)
points(Xa28,col="green")
```

```{r}
boxplot(t(Qdur),outpch=20, outcex=0.5)
points(Xdur,col="purple")
```

```{r}
boxplot(t(Qmhw),outpch=20, outcex=0.5)
points(Xmhw,col="orange")
```

Calculate exceedance events and durations

```{r, warning=FALSE}
library(reshape2)
library(ggplot2)
library(patchwork)
mcnt = matrix(0, nrow=500, ncol=4); lcnt=mcnt
dcnt=cnt*matrix(rep(1:50,500), byrow=TRUE, nrow=500)
for (i in 1:500) {
  n4=sum(cnt[i,1:4])
  n10=sum(cnt[i,5:10])
  n20=sum(cnt[i,11:20])
  n50=sum(cnt[i,21:50])
  mcnt[i,]=c(n4,n10,n20,n50)
  d4=sum(dcnt[i,1:4])
  d10=sum(dcnt[i,5:10])
  d20=sum(dcnt[i,11:20])
  d50=sum(dcnt[i,21:50])
  lcnt[i,]=c(d4,d10,d20,d50)
}
cdf=data.frame(mcnt/NY)
colnames(cdf)=c("D4","D10","D20","D50")
cmdf=melt(cdf, id=NULL)
ddf=data.frame(lcnt/NY)
colnames(ddf)=c("D4","D10","D20","D50")
dmdf=melt(ddf, id=NULL)
g1=ggplot(data=cmdf, aes(x=variable, y=value, color=variable)) +
  geom_boxplot() +
  ylab("Exceedance events/year")
g2=ggplot(data=dmdf, aes(x=variable, y=value, color=variable)) +
  geom_boxplot() +
  ylab("Exceedance days/year")
print(g1+g2)
#write.csv(cbind(cdf,ddf),"EvCntCV.csv")
```

**Extend simulations to 2100**

```{r}
NS=500  #number of simulations
NY=120 #number of years
NR=NY*365 #record length in days
Z=matrix(0,nrow=NR,ncol=NS)
decY=0:(NY*365)/365+1982; decY=decY[-(length(decY))]
YMD=date_decimal(decY); YMD=as.Date(YMD); M=month(YMD);
ZY=year(YMD)
YV=unique(ZY)

for (j in 1:NS){
  Er = rnorm(NR,mean=0,sd=sig_e)
  for (i in 2:NR) {Z[i,j] = slope*Z[i-1,j] + Er[i]}
}

Ztrend=fitmS$coefficients[1]+fitmS$coefficients[2]*decY
ZsV=rep(Xseas,NY)
Zt=Z+Ztrend  #add trend from SST data
Zst=Zt+ZsV   #add seasonal cycle to trend
ZthV=rep(Xthresh,NY)

```

Calculate MHWs

```{r}
Zexp=matrix(0,nrow=NY,ncol=NS); Za28=Zexp; Zint=Zexp; Zdur=Zexp; Zmhw=Zexp
frachi = rep(0,NS)
for (j in 1:NS) {
  Xhi=rep(0,NR); N28=Xhi
  Xhi[(Zst[,j]-ZthV)>0]=1
  frachi[j]=sum(Xhi)/NR
  dX=diff(Xhi)
  nd = which(dX==1)
  np = which(dX==-1)
  if (np[1]<nd[1]) np=np[-1]
  if (length(np) != length(nd)) nd=nd[-length(nd)]
  durXHi=np-nd
  n5=which(durXHi>=5)
  inten=rep(0,length(n5))
  for (k in 1:length(n5)){
    inten[k]=max(Zt[nd[n5[k]]:np[n5[k]],j])
  }
  Dint=aggregate(inten ~ ZY[nd[n5]], FUN=max, na.action=na.pass)
  colnames(Dint)=c("Y","intMHW")
  Sint=rep(0,NY)
  Sint[Dint$Y-YV[1]+1]=Dint$intMHW
  Zint[,j]=Sint
  
  Dexp=aggregate(durXHi[n5] ~ ZY[nd[n5]], FUN=sum, na.action=na.pass)
  colnames(Dexp)=c("Y","expMHW")
  Sexp=rep(0,NY)
  Sexp[Dexp$Y-YV[1]+1]=Dexp$expMHW
  Zexp[,j]=Sexp
  
  N28[Zst[,j]>=28]=1
  a28=aggregate(N28 ~ ZY, FUN=sum, na.action=na.pass)    
  colnames(a28)=c("Y","abv28")
  Sa28=rep(0,NY)
  Sa28[a28$Y-YV[1]+1]=a28$abv28
  Za28[,j]=Sa28
  
  Ddur=aggregate(durXHi[n5] ~ ZY[nd[n5]], FUN=mean, na.action=na.pass)
  colnames(Ddur)=c("Y","durMHW")
  Sdur=rep(0,NY)
  Sdur[Ddur$Y-YV[1]+1]=Ddur$durMHW
  Zdur[,j]=Sdur
  
  Dmhw=aggregate(nd[n5] ~ ZY[nd[n5]], FUN=NROW, na.action=na.pass)
  colnames(Dmhw)=c("Y","nMHW")
  Smhw=rep(0,NY)
  Smhw[Dmhw$Y-YV[1]+1]=Dmhw$nMHW
  Zmhw[,j]=Smhw
}
```


```{r}
boxplot(t(Zint),outpch=20, outcex=0.5)
points(Xint,col="blue")
```

```{r}
boxplot(t(Zexp),outpch=20, outcex=0.5)
points(Xexp,col="red")
```

```{r}
boxplot(t(Za28),outpch=20, outcex=0.5)
points(Xa28,col="green")
```

```{r}
boxplot(t(Zdur),outpch=20, outcex=0.5)
points(Xdur,col="purple")
```

```{r}
boxplot(t(Zmhw),outpch=20, outcex=0.5)
points(Xmhw,col="orange")
```


